---
title: "Distribution of Lakes Across Iowa Counties"
author: "Huy Nguyen, Nhat Le"
date: "2024-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Topic
Understanding the distribution of lakes across the 99 Iowa counties and examining potential correlations with agricultural land areas.

# Team Members
1. Huu Huy Nguyen

2. Nhat Le

# Introduction 
This project investigates the spatial distribution of 913 lakes in Iowa and explores the relationship between lake density and agricultural land in each county.

# Data Description
The project utilizes:

	1.	Lake Data: From data.iowa.gov, providing details about 913 lakes.
	
	2.	County Boundary Data: From the tigris library, offering spatial geometries and areas of Iowa counties.
	
	3.	Agricultural Land Data: Potentially sourced from the USDA for information on agricultural land areas by county.

# Methodology

### Libraries to be used

```{r}
library(tigris)
library(dplyr)
library(stringr)
library(tidyr)
library(sf)
```

### Data Preparation

```{r}
counties <- counties(state = "IA", cb = TRUE, class = "sf")
counties_area <- counties %>% mutate(county_area = st_area(geometry))
counties_area <- counties_area %>% select(NAME, county_area) %>% st_set_geometry(NULL)
```

```{r}
lakes <- read.csv('https://data.iowa.gov/resource/2mmi-txzx.csv')
```

For this part, we may collect a dataset about agricultural land in Iowa counties, provided by U.S. Department of Agriculture. We can also explore some spatial visualization of the data.

### Data Cleaning and Transformation
1. Standardize the names of counties between datasets.
	
2. Convert spatial data into sf format for analysis.

3. Ensure CRS (Coordinate Reference System) alignment for spatial operations.

### Data Visualization
Utilize R packages to visualize the spatial patterns of lake distribution.